---
title: "Hands-on Ex 01"
date: "Last updated on `r format(Sys.time(), '%d %B, %Y')`"
format:
  html:
    code-fold: false
    code-summary: "code block"
---

<font size = "5">**Geospatial Data Wrangling**</font>

# Installing R packages

```{r}
pacman::p_load(sf, tidyverse)
```

# Importing Geospatial Data

## Polygon Feature Data

```{r}
mpsz <- sf::st_read(
   dsn = "data/geospatial",
   layer = "MP14_SUBZONE_WEB_PL"
 )
```

::: {.solvebox .solve data-latex="solve"}

**st_read()** reveals that the geospatial objects are multipolygon features. There are a total of 323 multipolygon features and 15 fields in the **mpsz** simple feature data frame

:::

## Polyline Feature Data

```{r}
cyclingpath <- sf::st_read(
   dsn = "data/geospatial", 
   layer = "CyclingPathGazette"
  )
```

::: {.solvebox .solve data-latex="solve"}

There are a total of 2558 features and 2 fields in cyclingpath linestring feature data frame and it is in **svy21** projected coordinates system

:::

## GIS Data (KML format)

```{r}
preschool <- sf::st_read(
   "data/geospatial/PreSchoolsLocation.kml")
```

::: {.solvebox .solve data-latex="solve"}

This reveals that `preschool` is a **point feature** data frame. There are a total of 2290 features and 2 fields and is in **wgs84** coordinates system.

:::

# What is the Content of A Simple Feature Data Frame?

::: panel-tabset

## **st_geometry()**

```{r}
st_geometry(mpsz)
```
## **glimpse()**

This gives more information about the associated attributes in the data frame:

```{r}
glimpse(mpsz)
```

## **head()**

```{r}
head(mpsz)
```
:::

# Plotting the Geospatial Data

## Different variations of `plot()`

::: panel-tabset

## **simple plot()**

```{r}
#| warning: false
plot(mpsz)
```
## **geometry only**

```{r}
# Plot only the geometry using st_geometry

plot(st_geometry(mpsz))
```

## **specific attribute**

```{r}
plot(mpsz["SUBZONE_C"])
```

:::





